Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2025-05-24 manuscript/manuscript.tex
Shortening methods section

VA involves two major components: survey and COD assignment \citep{worldhealthorganizationVerbalAutopsyStandards2023,chandramohanEstimatingCausesDeath2021,worldhealthorganizationVerbalAutopsyStandards2007}. In the survey component, trained lay surveyors interview those familiar with the deceased (e.g. living spouse, children, family, friends) to gather information using standardized questionnaires and open narratives. In the COD assignment component, physicians evaluate information available from the questionnaires and open narratives to assign probable CODs. This component has been criticized to be difficult to reproduce due to reliance on physician assignment \citep{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005,mccormickProbabilisticCauseofDeathAssignment2016b,morrisFactorsAssociatedPhysician2010,solemanVerbalAutopsyCurrent2006}. As an alternative to physician assignment, computer models, such as InterVA \citep{byassIntegratedApproachProcessing2019} and InSilicoVA \citep{mccormickProbabilisticCauseofDeathAssignment2016b}, have been studied to automatically assign CODs with performances close to physicians at the population level, but poor performances at the individual level \citep{jhaAutomatedPhysicianAssignment2019,leitaoComparisonPhysiciancertifiedVerbal2014,desaiPerformanceFourComputercoded2014,tungaVerbalAutopsyModels2021,otiVerbalAutopsyInterpretation2010}. These computer models often utilize data from the structured questionnaire, but often omit the free-text open narrative, which misses latent information, such as chronology or health-seeking behaviors, that may potentially help models perform better than using the questionnaire alone \citep{jebleeAutomaticallyDeterminingCause2019,blancoExtractingCauseDeath2021,kingQualityDiagnosticValue2016}.

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text, such as question answering, code generation, and even medical diagnosis \citep{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. In 2022, a widely-available LLM called ChatGPT was released by OpenAI with capabilities of answering natural language text inquiries using training data up to September 2021. ChatGPT-3 was based on several Generative Pre-trained Transformer (GPT) models between 2018 to 2020, namely GPT-1 to GPT-3, which had notable differences in training data sizes of 5 gigabytes to 45 terabytes from web sources that resulted in 117 million to 175 billion parameter models \citep{wuBriefOverviewChatGPT2023}. In March 2023, ChatGPT-4 was released with human-level performance on various professional and academic exams and benchmarks that outperformed ChatGPT-3 \citep{openaiGPT4TechnicalReport2023}. Given the limited usage of free-text open narratives in computer models for determining CODs, and recent advances in LLMs that leverage natural language text prompts, we conduct a case study with Sierra Leone deaths from VA in 2019 to 2022 to compare four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physicians for determining CODs.

@@

Initially, 11,920 records from the HEAL-SL study \citep{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022} were collected from dual-coded EVA, where each record was randomly coded by two different physicians that assigned CODs as International Classification of Diseases Revision 10 (ICD-10) codes \citep{worldhealthorganizationICD10InternationalStatistical2011}. Physicians were able to assign CODs for 11,820 of the 11,920 records, where 100 of these records could not be assigned a COD due to missing or inadequate information (e.g. low quality narrative, data loss). For each record, two codes were assigned by two different randomly selected physicians, where codes were evaluated for agreement using Central Medical Evaluation Agreement 10 (CMEA-10) codes. CMEA-10 groups a range of similar ICD-10 codes together, where if they are in agreement if they are within the same group \citep{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 2). When codes were not in agreement, a record enters the reconciliation phase, where the two physicians were provided reasoning and initial codes from each other to: (1) keep their initial code (2) assign the other physician's code or (3) assign a new code. If codes were not in agreement after the reconciliation phase, a record enters the adjudication phase, where a third senior physician evaluates both physicians' reasoning and codes before and after reconciliation, and assigns a final code based on their evaluation.

The 11,820 physician coded records were further filtered for records where both physicians agreed on the assigned codes (records that were not reconciled or adjudicated) resulting in 6942 physician agreed records. Since computer models were compared to physicians in this study, there was more certainty that COD assignments agreed by both physicians were representative of physician assignment than when they disagreed \citep{barnettComparativeAccuracyDiagnosis2019,morrisFactorsAssociatedPhysician2010,hsiaoFactorsAssociatedPhysician2012}. The 6942 records were converted into CGHR-10 codes (see Additional File 1) that generalized ICD-10 codes into 19, 10, and 7 categories for the adult (12 to 69 years), child (28 days to 11 years), and neonatal (under 28 days) age groups.  After conversion, a final total of 6939 physician agreed records (3826 adult, 2636 child, and 477 neonatal) were used for modelling and performance evaluation, where three records were removed as their ICD-10 codes did not have a matching CGHR-10 code. The 6939 physician agreed records were evenly distributed in terms of sex and age, except for children aged 1-5 years (n=1633, 24\%), while Malaria (n=799, 21\%) was the COD for about a fifth of the records. See Appendix \ref{app-data} for further details on data distributions across age groups.

@@

Four computer models were used to assign COD for each of the 6939 physician agreed records: GPT-3.5, GPT-4, InterVA-5, and InSilicoVA. Each model required pre-processing of the 6939 records into input data, and standardization of output COD codes from models for performance evaluation as not all models produced comparable codes across outputs. Although each model can assign multiple CODs per record, only the first generated COD response from GPT-3.5 and GPT-4, and the most probable COD from InterVA-5 and InSilicoVA were used for evaluation. All model outputs were converted to CGHR-10 codes to evaluate performances of models for COD assignment relative to physicians.

\subsubsection{Overview of GPT-3.5/4, InterVA-5, and InSilicoVA}\label{methods-models-overview}

GPT-3.5 \citep{brownLanguageModelsAre2020} and GPT-4 \citep{openaiGPT4TechnicalReport2023} are LLMs that utilize deep neural networks with transformer architectures \citep{vaswaniAttentionAllYou2017} and reinforcement learning from human feedback \citep{ouyangTrainingLanguageModels2022,christianoDeepReinforcementLearning2017,stiennonLearningSummarizeHuman2020,wirthSurveyPreferencebasedReinforcement2017} to follow instructions from prompts and provide human-level responses, with known differences in GPT-4 possessing multimodal capabilites (e.g. image/voice input/output), more recent training data, and improved responses compared to ChatGPT-3 \citep{wuBriefOverviewChatGPT2023}. GPT-3.5 and GPT-4 were instructed with text prompts to assign CODs based on the open narrative. InterVA-5 and InSilicoVA are widely used and studied standard statistical models \citep{murrayUsingVerbalAutopsy2014,otiVerbalAutopsyInterpretation2010,benaraEvaluationMethodsAssigning2023,chandramohanEstimatingCausesDeath2021,tungaVerbalAutopsyModels2021,leitaoComparisonPhysiciancertifiedVerbal2014,jhaAutomatedPhysicianAssignment2019} for COD assignment in VAs under the openVA framework \citep{liOpenVAToolkitVerbal2023}. InterVA-5 applies Bayesian probabilistic modelling \citep{bayesEssaySolvingProblem1958} using a set of standardized symptoms from reports and related conditional probabilities from medical experts to assign CODs based on the highest probability \citep{byassIntegratedApproachProcessing2019,byassStrengtheningStandardisedInterpretation2012}. InSilicoVA improves upon InterVA (e.g. comparable probabilities across individuals, measures of uncertainty, and inclusion of additional data sources) with a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) simulations \citep{brooksMarkovChainMonte1998,chibMarkovChainMonte2001,hanMarkovChainMonte2001} to incorporate multiple sources of uncertainty for assigning CODs based on the highest probability \citep{mccormickProbabilisticCauseofDeathAssignment2016b}. For assigning CODs, GPT-3.5 and GPT-4 require prompts containing conversation-like textual instructions as input, while InterVA-5 and InSilicoVA require structured symptom and associated sociodemographic data as input.

\subsubsection{Input Data and Preprocessing}\label{methods-models-input}

\begin{lstlisting}
You are a physician with expertisein determining underlying causes of death in Sierra Leone by assigning the most probable ICD-10 code for each death using verbal autopsy narratives. Return only the ICD-10 code without description. E.g. A00. If there are multiple ICD-10 codes, show one code per line.
\end{lstlisting}

@@

The \verb|openVA| R package was used to run InterVA-5 and InSilicoVA models to assign CODs for each record in R version 4.3.1. The \verb|openVA| package version 1.1.1 used dependent packages \verb|InterVA5| version 1.1.3 and \verb|InSilicoVA| version 1.4.0. The \verb|Nsim| (number of iterations to run) parameter \citep{liInSilicoVAProbabilisticVerbal2022} for InSilicoVA was set to 9500, while the \verb|HIV| (level of prevalence of human immunodeficiency virus) and \verb|Malaria| (level of prevalence of Malaria) parameters \citep{thomasInterVA5ReplicateAnalyse2021} for InterVA-5 were both set to \verb|'h'| (high) reflecting HIV and Malaria disease assumptions in Sierra Leone \citep{yendewaHIVAIDSSierra2018,walkerMalariaMorbidityMortality2015}.

% Note that the default value of \verg|Nsim=10000| for InSilicoVA ran until 9500 iterations before it stopped due to errors, thus Nsim=9500 was used and ran successfully for all iterations.

\subsubsection{Output Data and Code Conversions}\label{methods-models-output}

@@

\begin{table}[h]
    \caption{Leading causes globally in 2021 and most relevant models.}\label{tab-leadcod}%
    \begin{tabular}{@{}llll@{}}
    \toprule
    \makecell{Leading Causes of Death\\($\sim$55\% of 68M deaths)\footnotemark[1]} & \makecell{Deaths\\(of 68M)\footnotemark[2]} & \makecell{Best\\Model(s)} & \makecell{PCCC} \\
    \midrule

    Ischaemic heart disease & 9M (13\%) & GPT-4 & 0.65 (n=168)\\ 
    Stroke & 7M (10\%) & GPT-4 & 0.77 (n=331)\\
    Cancers & 4.3M (4\%) & GPT-4 & 0.91 (n=49)\\
    Lower respiratory infections & 2.4M (3\%) & GPT-3.5/4 & 0.78 (n=180)\footnotemark[3]\\
    Diabetes mellitus & 1.6M (2\%) & GPT-3.5 & 0.92 (n=27)\\
    Kidney disease & 1.4M (2\%)  & GPT-3.5 & 0.92 (n=27)\footnotemark[4]\\
    Tuberculosis & 1.4M (2\%) & InSilicoVA & 0.71 (n=171)\\
    Hypertensive heart disease & 1.4M (2\%) & GPT-4 & 0.78 (n=107)\footnotemark[5]\\
    Cirrhosis of the liver & 1.3M (2\%) & GPT-3.5 & 0.78 (n=128)\footnotemark[6]\\ 
    Diarrhoeal diseases & 1.2M (2\%) & GPT-4 & 0.68 (n=205)\\ 
    Road injury & 1,183 (2\%) & InSilicoVA & 0.84 (n=357)\\
    Preterm birth complications & 0.9M (1\%) & GPT-4 & 0.71 (n=73)\\
    Falls & 0.7M (1\%) & GPT-4 & 0.89 (n=492)\footnotemark[7]\\

    \botrule
    \end{tabular}
    \footnotetext[1]{Does not include COVID-19, alzheimer disease, other dementias, and self-harm. Trachea, bronchus, lung, colon, rectum, stomache, and breast cancers were generalized into cancers.}
    \footnotetext[2]{Total of $\sim$68 Million (M) deaths globally. Numbers are rounded.}
    \footnotetext[3]{Mean of chronic and acute respiratory infections.}
    \footnotetext[4]{Derived from diabetes mellitus.}
    \footnotetext[5]{Derived from other cardiovascular diseases.}
    \footnotetext[6]{Derived from liver and alcohol related diseases.}
    \footnotetext[7]{Mean of adult and child injuries.}
\end{table}

@@

The 6939 physician agreed records were relatively evenly distributed (approximately 40-57\% male and female) for the adult, child, and neonatal age groups in terms of sex (Figure \ref{fig-data-agesex}) with the most records in the 1-5 year (n=1633, 24\%) age range, and the least records in the 10-14 year (n=135, 2\%) and 7-27 day (n=82, 1\%) age ranges (Figure \ref{fig-data-agerange}). The other 15 five-year age ranges within 0 days to 69 years were relatively evenly distributed with approximately 4-6\% of the 6939 records each. The 3826 adult records had most records in the 65-69 year (n=575, 15\%), and the least in the 10-14 year (n=135, 4\%) age ranges, while the other 10 age ranges were relatively evenly distributed at 7-9\% of all records (Figure \ref{fig-data-agerange-adult}). For adult CODs, Malaria (n=799, 21\%) had the highest number of adult records, and cancers (n=49, 1\%), diabetes mellitus (n=27, \textless 1\%), and suicide (n=3, \textless 1\%) had the lowest number adult of adult records, while the other 15 CODs had between 2-11\% of all adult records (Figure \ref{fig-data-cod-adult}). The 2636 child records had most records in the 1-5 year (n=1633, 62\%) age range while the three other age ranges were relatively evenly distributed at 12-14\% of all records (Figure \ref{fig-data-agerange-child}). For child CODs, Malaria (n=1382, 52\%) and other infections (n=667, 25\%) had the highest number of child records, while nutritional deficiencies (n=11, \textless 1\%) and congenital anomalies (n=1, \textless 1\%) had the lowest number of records with the five other CODs between 3-7\% of all child records. The 477 neonatal records had two age ranges with most records in the 0-6 day (n=395, 83\%) age range and the rest in the 7-27 day (n=82, 17\%) age range (Figure \ref{fig-data-agerange-neo}). For neonatal CODs, stillbirth (n=172, 36\%) had the highest number of neonatal records, and ill-defined (n=23, 5\%), other (n=5, 1\%), and congenital anomalies (n=2, \textless 1\%) had the least number of records with the rest of the other three CODs between 15-22\% of all neonatal records (Figure \ref{fig-data-cod-neo}).

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{fig-map.pdf}
\caption{Case study data sampling areas.}\label{fig-map}
\end{figure}

\begin{sidewaystable}
\caption{Case study data by cause of death.}\label{tab-data-cod}
\begin{tabular*}{\textheight}{@{\extracolsep\fill}lllll}
    \toprule%
    Age Group  &  Cause of Death (COD)  &  Female  &  Male  &  Total  \\
    \midrule
    \makecell[l]{\\Adult (n=3826, 55.1\%, 18 CODs)\\Adult Female (n=1681, 43.9\%)\\Adult Male (n=2145, 56.1\%)}  &  Acute Respiratory Infections  &  48 (45.7\%)  &  57 (54.3\%)  &  105 (2.7\%)  \\ 
        &  Cancers  &  32 (65.3\%)  &  17 (34.7\%)  &  49 (1.3\%)  \\ 
        &  Chronic Respiratory Diseases  &  29 (38.7\%)  &  46 (61.3\%)  &  75 (2\%)  \\ 
        &  Diabetes Mellitus  &  14 (51.9\%)  &  13 (48.1\%)  &  27 (0.7\%)  \\ 
        &  Diarrhoeal Diseases  &  102 (49.8\%)  &  103 (50.2\%)  &  205 (5.4\%)  \\ 
        &  Ill-Defined  &  56 (47.9\%)  &  61 (52.1\%)  &  117 (3.1\%)  \\ 
        &  Ischemic Heart Disease  &  89 (53\%)  &  79 (47\%)  &  168 (4.4\%)  \\ 
        &  Liver And Alcohol Related Diseases  &  58 (45.3\%)  &  70 (54.7\%)  &  128 (3.3\%)  \\ 
        &  Malaria  &  372 (46.6\%)  &  427 (53.4\%)  &  799 (20.9\%)  \\ 
        &  Maternal Conditions  &  130 (100\%)  &  N/A  &  130 (3.4\%)  \\ 
        &  Other Cardiovascular Diseases  &  59 (55.1\%)  &  48 (44.9\%)  &  107 (2.8\%)  \\ 
        &  Other Noncommunicable Diseases  &  160 (38.6\%)  &  254 (61.4\%)  &  414 (10.8\%)  \\ 
        &  Other Injuries  &  83 (23.2\%)  &  274 (76.8\%)  &  357 (9.3\%)  \\ 
        &  Road And Transport Injuries  &  73 (29.1\%)  &  178 (70.9\%)  &  251 (6.6\%)  \\ 
        &  Stroke  &  147 (44.4\%)  &  184 (55.6\%)  &  331 (8.7\%)  \\ 
        &  Suicide  &  N/A  &  3 (100\%)  &  3 (0.1\%)  \\ 
        &  Tuberculosis  &  54 (31.6\%)  &  117 (68.4\%)  &  171 (4.5\%)  \\ 
        &  Unspecified Infections  &  175 (45\%)  &  214 (55\%)  &  389 (10.2\%)  \\ 
    \makecell[l]{\\Child (n=2636, 38\%, 9 CODs)\\Child Female (n=1290, 48.9\%)\\Child Male (n=1346, 51.1\%)}  &  Congenital Anomalies  &  1 (100\%)  &  N/A  &  1 (0\%)  \\ 
        &  Diarrhoeal Diseases  &  79 (45.1\%)  &  96 (54.9\%)  &  175 (6.6\%)  \\ 
        &  \makecell[l]{Epilepsy, Leukaemia, And\\Other Noncommunicable Diseases}  &  61 (53.5\%)  &  53 (46.5\%)  &  114 (4.3\%)  \\ 
        &  Ill-Defined  &  34 (48.6\%)  &  36 (51.4\%)  &  70 (2.7\%)  \\ 
        &  Injuries  &  51 (37.8\%)  &  84 (62.2\%)  &  135 (5.1\%)  \\ 
        &  Malaria  &  680 (49.2\%)  &  702 (50.8\%)  &  1382 (52.4\%)  \\ 
        &  Nutritional Deficiencies  &  7 (63.6\%)  &  4 (36.4\%)  &  11 (0.4\%)  \\ 
        &  Other Infections  &  338 (50.7\%)  &  329 (49.3\%)  &  667 (25.3\%)  \\ 
        &  Pneumonia  &  39 (48.1\%)  &  42 (51.9\%)  &  81 (3.1\%)  \\ 
    \makecell[l]{\\Neonate (n=477, 6.9\%, 7 CODs)\\Neonate Female (n=227, 47.6\%)\\Neonate Male (n=250, 52.4\%)}  &  Birth Asphyxia And Birth Trauma  &  38 (36.9\%)  &  65 (63.1\%)  &  103 (21.6\%)  \\ 
        &  Congenital Anomalies  &  2 (100\%)  &  N/A  &  2 (0.4\%)  \\ 
        &  Ill-Defined  &  11 (47.8\%)  &  12 (52.2\%)  &  23 (4.8\%)  \\ 
        &  Neonatal Infections  &  49 (49.5\%)  &  50 (50.5\%)  &  99 (20.8\%)  \\ 
        &  Other  &  2 (40\%)  &  3 (60\%)  &  5 (1\%)  \\ 
        &  Prematurity And Low Birthweight  &  39 (53.4\%)  &  34 (46.6\%)  &  73 (15.3\%)  \\ 
        &  Stillbirth  &  86 (50\%)  &  86 (50\%)  &  172 (36.1\%)  \\
    \botrule
\end{tabular*}
\end{sidewaystable}

\begin{table}
\caption{Case study data by age range.}\label{tab-data-age}
\begin{tabular}{@{}lllll@{}}
\toprule%
    Age Group  &  Age Range  &  Female  &  Male  &  Total  \\
    \midrule
        \makecell[l]{\\Adult (n=3826, 55.1\%, 18 CODs)\\Adult Female (n=1681, 43.9\%)\\Adult Male (n=2145, 56.1\%)}  &  10-14 Years  &  51 (37.8\%)  &  84 (62.2\%)  &  135 (3.5\%)  \\ 
        &  15-19 Years  &  115 (42.8\%)  &  154 (57.2\%)  &  269 (7\%)  \\ 
        &  20-24 Years  &  146 (53.1\%)  &  129 (46.9\%)  &  275 (7.2\%)  \\ 
        &  25-29 Years  &  159 (45.2\%)  &  193 (54.8\%)  &  352 (9.2\%)  \\ 
        &  30-34 Years  &  174 (50.9\%)  &  168 (49.1\%)  &  342 (8.9\%)  \\ 
        &  35-39 Years  &  153 (45.4\%)  &  184 (54.6\%)  &  337 (8.8\%)  \\ 
        &  40-44 Years  &  134 (42\%)  &  185 (58\%)  &  319 (8.3\%)  \\ 
        &  45-49 Years  &  148 (47\%)  &  167 (53\%)  &  315 (8.2\%)  \\ 
        &  50-54 Years  &  134 (39.6\%)  &  204 (60.4\%)  &  338 (8.8\%)  \\ 
        &  55-59 Years  &  96 (37.6\%)  &  159 (62.4\%)  &  255 (6.7\%)  \\ 
        &  60-64 Years  &  128 (40.8\%)  &  186 (59.2\%)  &  314 (8.2\%)  \\ 
        &  65-69 Years  &  243 (42.3\%)  &  332 (57.7\%)  &  575 (15\%)  \\ 
        \makecell[l]{\\Child (n=2636, 38\%, 9 CODs)\\Child Female (n=1290, 48.9\%)\\Child Male (n=1346, 51.1\%)}  &  1-5 Months  &  146 (47.4\%)  &  162 (52.6\%)  &  308 (11.7\%)  \\ 
        &  6-11 Months  &  160 (50.8\%)  &  155 (49.2\%)  &  315 (11.9\%)  \\ 
        &  1-5 Years  &  822 (50.3\%)  &  811 (49.7\%)  &  1633 (61.9\%)  \\ 
        &  6-11 Years  &  162 (42.6\%)  &  218 (57.4\%)  &  380 (14.4\%)  \\ 
        \makecell[l]{\\Neonate (n=477, 6.9\%, 7 CODs)\\Neonate Female (n=227, 47.6\%)\\Neonate Male (n=250, 52.4\%)}  &  0-6 Days  &  184 (46.6\%)  &  211 (53.4\%)  &  395 (82.8\%)  \\ 
        &  7-27 Days  &  43 (52.4\%)  &  39 (47.6\%)  &  82 (17.2\%)  \\
    \botrule
\end{tabular}
\end{table}

@@

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{fig-methods.pdf}
\caption{Case study methods.}\label{fig-methods}
\end{figure}

\subsection{Cause Specific Mortality Fraction (CSMF) Accuracy}\label{methods-eval-csmfa}

>>>>

VA involves two major components: survey and COD assignment \citep{worldhealthorganizationVerbalAutopsyStandards2023,chandramohanEstimatingCausesDeath2021,worldhealthorganizationVerbalAutopsyStandards2007}. In the survey component, trained lay surveyors interview those familiar with the deceased (e.g. living spouse, children, family, friends) to gather information using standardized questionnaires and open narratives. In the COD assignment component, physicians evaluate information available from the questionnaires and open narratives to assign probable CODs. This component has been criticized to be difficult to reproduce due to reliance on physician assignment \citep{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005,mccormickProbabilisticCauseofDeathAssignment2016b,morrisFactorsAssociatedPhysician2010,solemanVerbalAutopsyCurrent2006}. As an alternative to physician assignment, computer models, such as InterVA \citep{byassIntegratedApproachProcessing2019} and InSilicoVA \citep{mccormickProbabilisticCauseofDeathAssignment2016b}, have been studied to automatically assign CODs with performances close to physicians at the population level, but poor performances at the individual level \citep{jhaAutomatedPhysicianAssignment2019,leitaoComparisonPhysiciancertifiedVerbal2014,desaiPerformanceFourComputercoded2014,tungaVerbalAutopsyModels2021,otiVerbalAutopsyInterpretation2010}. InterVA-5 and InSilicoVA are widely used and studied standard statistical models \citep{murrayUsingVerbalAutopsy2014,otiVerbalAutopsyInterpretation2010,benaraEvaluationMethodsAssigning2023,chandramohanEstimatingCausesDeath2021,tungaVerbalAutopsyModels2021,leitaoComparisonPhysiciancertifiedVerbal2014,jhaAutomatedPhysicianAssignment2019} for COD assignment in VAs under the openVA framework \citep{liOpenVAToolkitVerbal2023}. InterVA-5 applies Bayesian probabilistic modelling \citep{bayesEssaySolvingProblem1958} using a set of standardized symptoms from reports and related conditional probabilities from medical experts to assign CODs based on the highest probability \citep{byassIntegratedApproachProcessing2019,byassStrengtheningStandardisedInterpretation2012}. InSilicoVA improves upon InterVA (e.g. comparable probabilities across individuals, measures of uncertainty, and inclusion of additional data sources) with a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) simulations \citep{brooksMarkovChainMonte1998,chibMarkovChainMonte2001,hanMarkovChainMonte2001} to incorporate multiple sources of uncertainty for assigning CODs based on the highest probability \citep{mccormickProbabilisticCauseofDeathAssignment2016b}. These computer models often utilize data from the structured questionnaire, but often omit the free-text open narrative, which misses latent information, such as chronology or health-seeking behaviors, that may potentially help models perform better than using the questionnaire alone \citep{jebleeAutomaticallyDeterminingCause2019,blancoExtractingCauseDeath2021,kingQualityDiagnosticValue2016}.

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text, such as question answering, code generation, and even medical diagnosis \citep{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. In 2022, a widely-available LLM called ChatGPT was released by OpenAI with capabilities of answering natural language text inquiries using training data up to September 2021. ChatGPT-3 was based on several Generative Pre-trained Transformer (GPT) models between 2018 to 2020, namely GPT-1 to GPT-3, which had notable differences in training data sizes of 5 gigabytes to 45 terabytes from web sources that resulted in 117 million to 175 billion parameter models \citep{wuBriefOverviewChatGPT2023}. In March 2023, ChatGPT-4 was released with human-level performance on various professional and academic exams and benchmarks that outperformed ChatGPT-3 \citep{openaiGPT4TechnicalReport2023}. GPT-3.5 \citep{brownLanguageModelsAre2020} and GPT-4 \citep{openaiGPT4TechnicalReport2023} are LLMs that utilize deep neural networks with transformer architectures \citep{vaswaniAttentionAllYou2017} and reinforcement learning from human feedback \citep{ouyangTrainingLanguageModels2022,christianoDeepReinforcementLearning2017,stiennonLearningSummarizeHuman2020,wirthSurveyPreferencebasedReinforcement2017} to follow instructions from prompts and provide human-level responses, with known differences in GPT-4 possessing multimodal capabilites (e.g. image/voice input/output), more recent training data, and improved responses compared to ChatGPT-3 \citep{wuBriefOverviewChatGPT2023}. Given the limited usage of free-text open narratives in computer models for determining CODs, and recent advances in LLMs that leverage natural language text prompts, we conducted a case study with Sierra Leone deaths from VA in 2019 to 2022 to compare four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physicians for determining CODs.

@@

Initially, 11,920 records from the HEAL-SL study \citep{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022} were collected from dual-coded EVA, where each record was randomly coded by two different physicians that assigned CODs as International Classification of Diseases Revision 10 (ICD-10) codes \citep{worldhealthorganizationICD10InternationalStatistical2011}. For each record, two codes were assigned by two different randomly selected physicians, where codes were evaluated for agreement using Central Medical Evaluation Agreement 10 (CMEA-10) codes. CMEA-10 groups a range of similar ICD-10 codes together, where if they are in agreement if they are within the same group \citep{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 2). When codes were not in agreement, a record enters the reconciliation phase, where the two physicians were provided reasoning and initial codes from each other to: (1) keep their initial code (2) assign the other physician's code or (3) assign a new code. If codes were not in agreement after the reconciliation phase, a record enters the adjudication phase, where a third senior physician evaluates both physicians' reasoning and codes before and after reconciliation, and assigns a final code based on their evaluation.

Since computer models were compared to physicians in this study, there was more certainty that COD assignments agreed by both physicians were representative of physician assignment than when they disagreed \citep{barnettComparativeAccuracyDiagnosis2019,morrisFactorsAssociatedPhysician2010,hsiaoFactorsAssociatedPhysician2012}. Thus, we used 6942 physician agreed records of the 11,920 total records. For better comparison, we standardized all codes to CGHR-10 codes (see Additional File 1) that generalized ICD-10 codes into 19, 10, and 7 categories for the adult (12 to 69 years), child (28 days to 11 years), and neonatal (under 28 days) age groups. After conversion, a final total of 6939 physician agreed records (3826 adult, 2636 child, and 477 neonatal) were used for modelling and performance evaluation. See Appendix \ref{app-methods-data} and \ref{app-data} for further details on data preprocessing and characteristics of the physician agreed records.

@@

Four computer models were used to assign COD for each of the 6939 physician agreed records: GPT-3.5, GPT-4, InterVA-5, and InSilicoVA. For GPT-3.5 and GPT-4, the following user prompt was used to instruct each model to produce COD assignments as ICD-10 codes, where \verb|<age>| and \verb|<sex>| from the questionnaire, and \verb|<narrative>| from the narratives, were replaced with values from the data:

@@

\noindent For InterVA-5 and InSilicoVA, the standardized questionnaire data from EVA were converted into into OpenVA format \citep{liOpenVAToolkitVerbal2023}, and fed into each model to produce COD assignments as WHOVA 2016 codes. All model outputs (ICD-10 for GPT-3.5/4 and WHOVA-2016 for InSilicoVA/InterVA-5) were converted to CGHR-10 codes to evaluate performances of models for COD assignment relative to physicians. See Appendix \ref{app-methods-model} for additional details regarding input parameters and code conversions for each model.

@@

\begin{table}[h]
    \caption{Leading causes globally in 2021 and most relevant models.}\label{tab-leadcod}%
    \begin{tabular}{@{}llll@{}}
    \toprule
    \makecell{Leading Causes of Death\\($\sim$53\% of 68M deaths)\footnotemark[1]} & \makecell{Deaths\\(\% of 68M)\footnotemark[2]} & \makecell{Best\\Model(s)} & \makecell{PCCC} \\
    \midrule

    Ischaemic heart disease & 9M (13\%) & GPT-4 & 0.65 (n=168)\\ 
    Stroke & 7M (10\%) & GPT-4 & 0.77 (n=331)\\
    Cancers & 4.3M (4\%) & GPT-4 & 0.91 (n=49)\\
    Lower respiratory infections & 2.4M (3\%) & GPT-3.5/4 & 0.78 (n=180)\footnotemark[3]\\
    Diabetes mellitus & 1.6M (2\%) & GPT-3.5 & 0.92 (n=27)\\
    Tuberculosis & 1.4M (2\%) & InSilicoVA & 0.71 (n=171)\\
    Hypertensive heart disease & 1.4M (2\%) & GPT-4 & 0.78 (n=107)\footnotemark[4]\\
    Cirrhosis of the liver & 1.3M (2\%) & GPT-3.5 & 0.78 (n=128)\footnotemark[5]\\ 
    Diarrhoeal diseases & 1.2M (2\%) & GPT-4 & 0.68 (n=205)\\ 
    Road injury & 1,183 (2\%) & InSilicoVA & 0.84 (n=357)\\
    Preterm birth complications & 0.9M (1\%) & GPT-4 & 0.71 (n=73)\\
    Falls & 0.7M (1\%) & GPT-4 & 0.89 (n=492)\footnotemark[6]\\

    \botrule
    \end{tabular}
    \footnotetext[1]{COVID-19, kidney disease, alzheimer disease, other dementias, and self-harm were excluded as a relevant CGHR-10 code was not present. Trachea, bronchus, lung, colon, rectum, stomache, and breast cancers were generalized into cancers.}
    \footnotetext[2]{Percentage of $\sim$68 Million (M) deaths globally. Numbers are rounded.}
    \footnotetext[3]{Mean of chronic and acute respiratory infections.}
    \footnotetext[4]{Derived from other cardiovascular diseases.}
    \footnotetext[5]{Derived from liver and alcohol related diseases.}
    \footnotetext[6]{Mean of adult and child injuries.}
\end{table}

@@

@@

This section provides additional details on the methods described in Section \ref{methods}. An overview of the methods used in this study is seen in Figure \ref{fig-methods} as a five-step process. Section \ref{app-methods-data} provides details on the preprocessed data used for modelling. Section \ref{app-methods-model} describes the data and parameter inputs and outputs for each model, while Section \ref{app-methods-eval} details the evaluation of model outputs at the individual and population level across different CODs, age groups, and age ranges.

\begin{figure}[H]
\centering
\includegraphics[width=.7\textwidth]{fig-methods.pdf}
\caption{Case study methods.}\label{fig-methods}
\end{figure}

\subsection{CGHR-10 Physician Agreed Records}\label{app-methods-data}

Initially, 11,920 records were collected from dual-coded EVA in the HEAL-SL study. Physicians were able to assign CODs for 11,820 of the 11,920 records, where 100 of these records could not be assigned a COD due to missing or inadequate information (e.g. low quality narrative, data loss). The 11,820 physician coded records were further filtered for records where both physicians agreed on the assigned codes (records that were not reconciled or adjudicated) resulting in 6942 physician agreed records (based on comparisons using CMEA-10 codes, see Additional File 2). The 6942 records were converted into CGHR-10 codes (see Additional File 1) that generalized ICD-10 codes into 19, 10, and 7 categories for the adult (12 to 69 years), child (28 days to 11 years), and neonatal (under 28 days) age groups. After conversion, a final total of 6939 physician agreed records (3826 adult, 2636 child, and 477 neonatal) were used for modelling and performance evaluation, where three records were removed as their ICD-10 codes did not have a matching CGHR-10 code.

\subsection{Modelling Details}\label{app-methods-model}

Each model (GPT-3.5, GPT-4, InSilicoVA, and InterVA-5) required pre-processing of the 6939 records into input data, and standardization of output COD codes from models for performance evaluation as not all models produced comparable codes across outputs. Although each model can assign multiple CODs per record, only the first generated COD response from GPT-3.5 and GPT-4, and the most probable COD from InterVA-5 and InSilicoVA were used for evaluation. Section \ref{app-methods-input} describes the input data and parameters for each model, while Section \ref{app-methods-output} details the outputs from running each model.

\subsubsection{Input Data and Preprocessing}\label{app-methods-input}

\begin{lstlisting}
You are a physician with expertise in determining underlying causes of death in Sierra Leone by assigning the most probable ICD-10 code for each death using verbal autopsy narratives. Return only the ICD-10 code without description. E.g. A00. If there are multiple ICD-10 codes, show one code per line.
\end{lstlisting}
